JavaScript as an Embedded DSL

### 2012-03-21 papers/ecoop2012/article.tex
minor typos from official feedback

There are three widely known approaches for addressing the challenges outlined above. One is to create a standalone language or DSL that is compiled to JavaScript and provides different abstractions compared to JavaScript. Examples include WebDSL~\cite{webdsl}, Links~\cite{links,links-form} and Dart~\cite{dart}. However, this approach usually requires a lot effort in terms of language and compiler design, and tooling support, although WebDSL leverages Spoofax~\cite{spoofax} to alleviate this effort. Furthermore, it is not always clear how these languages interact with other languages on the server-side or with the existing JavaScript ecosystem on the client-side.

@@

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses an extension of Scala,
Scala-Virtualized~\cite{scala-virtualized}, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementation can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

@@

The \code{EffectExp} trait is part of the core LMS framework. It
inherits from \code{BaseExp} which instantiates \code{Rep} as
\code{Exp}. \code{Exp} represents an IR via two subclasses:
\code{Const} for constants, \code{Sym} for named values defining a
\code{Def}. \code{Def} is the base class for all IR nodes. In our
\code{DebugExp} trait, we extend \code{Def} to support a new IR node:
\code{Log}.

@@

\subsection{Typed APIs}\label{sec:typedApis}
First, we show how to incorporate an external JavaScript API in a
fully-typed way into our DSL. As an example, consider the following
DSL snippet, which gets the context of an HTML5 canvas element
selected by id:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].geContext()
\end{lstlisting}

@@

Notice that \code{document} has type \code{Rep[Element]}, and needs to
implement the interface of \code{ElementOps}, so that
\code{document.getElementById("canvas")} is well-typed. We achieve
this using an implicit conversion from \code{Rep[Element]} to
\code{ElementOps}.  At the DSL implementation level, the \code{ElementOps}
returned by this implicit conversion needs to generate an IR node for
each method call, as shown in the walkthrough
in~\ref{sec:defineDslComponent}. For example,
\code{document.getElementById("canvas")} becomes the IR node {\tt\small
  MethodCall(document, "getElementById", List("canvas"))}. This is a
mechanical transformation, implemented by \code{repProxy} using
reflection to intercept method calls and generate IR nodes based on
the method name and the arguments of the invocation. Note that this
use of reflection is purely at staging time, so there is no overhead
in the generated code.
\begin{lstlisting}
trait DomLift extends Dom with JSProxyBase {
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
  implicit def repToCanvasOps(x: Rep[Canvas]): CanvasOps =
    repProxy[Canvas,CanvasOps](x)
  implicit def repToContextOps(x: Rep[Context]): ContextOps =
    repProxy[Context,ContextOps](x)
}
\end{lstlisting}

Note also that since \code{getElementById} returns an arbitrary DOM
element, we need to cast it to a \code{Canvas} using
\code{as[Canvas]}. \code{as} is implemented simply as a cast in the
host language (no IR node is created):
\begin{lstlisting}
  trait AsRep {
    def as[T]: Rep[T]
  }
  implicit def asRep(x: Rep[_]): AsRep = new AsRep {
    def as[T]: Rep[T] = x.asInstanceOf[Rep[T]]
  }
\end{lstlisting}
Instead of this no-op implementation, it is possible to insert
run-time check-cast assertions in the generated JavaScript code.

@@

This possibility to escape into dynamic typing is particularly useful
in simplifying the incorporation of external JavaScript APIs and
libraries. Sometimes, the user might not want to build a statically
typed API for each external JavaScript library. In addition, for
some library, it might awkward to come up with such a statically-typed
interface. In general, we expect users to start with a dynamic API for
an external library, and progressively migrate it to a typed API as
the code matures. Consider again the example introduced in the typed
API section:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].getContext()
\end{lstlisting}

In a fully dynamic scenario, we declare the DOM API simply as:
\begin{lstlisting}
trait Dom extends JSDynamic {
  val document: DynamicRep
}
\end{lstlisting}
The \code{as[Canvas]} cast is not necessary in this dynamically typed
setting:
\begin{lstlisting}[escapechar=+]
val context = document.getElementById("canvas").geContext()
\end{lstlisting}

As a first step towards statically typing the API, we declare the type \code{Element}:
\begin{lstlisting}
trait Dom extends JSProxyBase with JSDynamic {
  val document: Rep[Element]
  trait Element
  trait ElementOps {
    def getElementById(id: Rep[String]): DynamicRep
  }
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
}
\end{lstlisting}
Since the method \code{getElementById} returns a \code{DynamicRep},
only the emphasized part of the expression is statically typed:
\begin{lstlisting}[escapechar=+]
val context = +{\it document.getElementById("canvas")}+.geContext()
\end{lstlisting}

@@

First-class functions are widely used in JavaScript. One particular
common case is for callback-driven programming. Therefore, staged
functions are important to interface with existingn libraries. They
will also play a crucial role in section~\ref{sec:cps}, where we
abstract over callback-driven programming.

@@

As for implementation, notice that the type of a \code{new JSLiteral
  \{...\}} expression is not \code{JSLiteral \{...\}} but
\code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
Scala-Virtualized compiler. JSLiteral inherits from a special marker
trait, which indicates its \code{new} expressions should be
reified. So the \code{new} expression is turned into a method call,
with information about all the field definitions. A complication is
that a field definition might reference another field being defined
(such as \code{b} being defined in terms of \code{a} in the example
above). So each definition is represented by its name and a {\it
  function} which takes a self type for the object literal. These
definition functions are evaluated in an order which allows the self
references to be resolved.

@@

Note that JavaScript version takes a callback as second argument that will be used to process the fetched tweets. We provide the rest of the logic that iterates over array of users and makes Ajax requests (listings~\ref{code:js:twitter_example}~\&~\ref{code:twitter_example}).

@@

\subsection{CPS and Staging}
Let's write the example from listing \ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
\begin{lstlisting}
def sleep(delay: Rep[Int]) = shift { k: (Rep[Unit]=>Rep[Unit]) =>
  window.setTimeout(fun(k), delay)
}
\end{lstlisting}

@@

\subsection{CPS for Interruptible Traversals}
We need to be able to interrupt our execution while traversing an array in order to implement functionality from listing \ref{code:twitter_example}. Let's consider a simplified example where we want to iterate over an array and sleep during each iteration. We present both code written in JavaScript and our DSL that achieves that (listings~\ref{code:js:sleep_iter}~\&~\ref{code:sleep_iter}).
\begin{figure}
\lstset{basicstyle=\ttfamily\scriptsize}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\begin{lstlisting}[caption=JavaScript,label=code:js:sleep_iter]
var xs = [1, 2, 3]
var i = 0
var msg = null
function f1() {
  if (i < xs.length) {
    window.setTimeout(f2, xs[i]*1000)
    msg = xs[i]
    i++
  }
}
function f2() {
  console.log(msg)
  f1()
}
f1()
\end{lstlisting}
\begin{lstlisting}[escapechar=|,caption=DSL,label=code:sleep_iter]
val xs = array(1, 2, 3)
// shorthand for xs.suspendable.foreach
for (x <- xs.suspendable) {
  sleep(x * 1000)
  console.log(String.valueOf(x))
}
log("done")

@@

Both programs, when executed, will print following to the JavaScript console:
\begin{lstlisting}
//pause for 1s
1
//pause for 2s
2
//pause for 3s
3
done
\end{lstlisting}

@@

The Ajax component completes the functionality required to run the Twitter example with one caveat:
The outer loop in listing \ref{code:twitter_example} uses \code{parSuspendable} to traverse arrays instead of the \code{suspendable} traversal variant we have defined in listing \ref{code:suspendable_foreach}.

@@

A cell object allows us to track dependencies between values. Whenever the \code{get} method is called and the value is not in the cell yet, the continuation will be added to a queue so it can be suspended until the value arrives. The \code{set} method takes care of resuming queued continuations. We expose \code{Cell} as external library using our typed API mechanism and we use it for implementing an abstraction for futures.

@@

We've implemented our DSL in Scala~\footnote{The code can be found at
  \url{http://github.com/namin/lms-sandbox}.}. We used our DSL to
develop a few web applications, which are simple but not
trivial. First, we implemented a few drawing examples like the
snowflakes of figure~\ref{fig:snowflakes}. We extended the Twitter
example from section~\ref{sec:cps}, which presents the latest tweets
from selected users in an interactive way. In order to do so, we
incorporated a useful subset of the DOM API and jQuery library using our
typed APIs.

\begin{wrapfigure}{r}{5cm}
\includegraphics[scale=0.2]{drawing}
\end{wrapfigure}
Finally, we developed a collaborative drawing application, which
includes a server-side component (implemented using the Jetty web
server). We use web sockets to communicate between the server and
clients. Each client transmits drawing commands to the server, which
broadcasts them to all the clients. When a new client joins, the
server sends the complete drawing history to the new client, and the
client reconstructs the image by playing the commands. A very
simple improvement is to make the server execute the drawing
commands as well, and keep an up-to-date bitmap of the drawing -- this
can easily be achieved by using the trivial embedding described in
section~\ref{sec:trivial-embedding}. New clients then just obtain the
bitmap instead of replaying the history, which is potentially be large.

>>>>

There are three widely known approaches for addressing the challenges outlined above. One is to create a standalone language or DSL that is compiled to JavaScript and provides different abstractions compared to JavaScript. Examples include WebDSL~\cite{webdsl}, Links~\cite{links,links-form} and Dart~\cite{dart}. However, this approach usually requires a lot of effort in terms of language and compiler design, and tooling support, although WebDSL leverages Spoofax~\cite{spoofax} to alleviate this effort. Furthermore, it is not always clear how these languages interact with other languages on the server-side or with the existing JavaScript ecosystem on the client-side.

@@

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses an extension of Scala,
Scala-Virtualized~\cite{scala-virtualized}, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementations can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

@@

The \code{EffectExp} trait is part of the core LMS framework. It
inherits from \code{BaseExp} which instantiates \code{Rep} as
\code{Exp}. \code{Exp} represents an IR via two subclasses:
\code{Const} for constants and \code{Sym} for named values defining a
\code{Def}. \code{Def} is the base class for all IR nodes. In our
\code{DebugExp} trait, we extend \code{Def} to support a new IR node:
\code{Log}.

@@

\subsection{Typed APIs}\label{sec:typedApis}
First, we show how to incorporate an external JavaScript API in a
fully-typed way into our DSL. As an example, consider the following
DSL snippet, which gets the context of an HTML5 canvas element
selected by id:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].getContext()
\end{lstlisting}

@@

Notice that \code{document} has type \code{Rep[Element]}, and needs to
implement the interface of \code{ElementOps}, so that
\code{document.getElementById("canvas")} is well-typed. We achieve
this using an implicit conversion from \code{Rep[Element]} to
\code{ElementOps}.  At the DSL implementation level, the \code{ElementOps}
returned by this implicit conversion needs to generate an IR node for
each method call, as shown in the walkthrough
in section~\ref{sec:defineDslComponent}. For example,
\code{document.getElementById("canvas")} becomes the IR node {\tt\small
  MethodCall(document, "getElementById", List("canvas"))}. This is a
mechanical transformation, implemented by \code{repProxy} using
reflection to intercept method calls and generate IR nodes based on
the method name and the arguments of the invocation. Note that this
use of reflection is purely at staging time, so there is no overhead
in the generated code.
\begin{lstlisting}
trait DomLift extends Dom with JSProxyBase {
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
  implicit def repToCanvasOps(x: Rep[Canvas]): CanvasOps =
    repProxy[Canvas,CanvasOps](x)
  implicit def repToContextOps(x: Rep[Context]): ContextOps =
    repProxy[Context,ContextOps](x)
}
\end{lstlisting}

Note also that since \code{getElementById} returns an arbitrary DOM
element, we need to cast it to a \code{Canvas} using
\code{as[Canvas]}. The \code{as} operation is implemented simply as a cast in the
host language (no IR node is created):
\begin{lstlisting}
  trait AsRep {
    def as[T]: Rep[T]
  }
  implicit def asRep(x: Rep[_]): AsRep = new AsRep {
    def as[T]: Rep[T] = x.asInstanceOf[Rep[T]]
  }
\end{lstlisting}
Instead of this no-op implementation, it is possible to insert
run-time check-cast assertions in the generated JavaScript code.

@@

This possibility to escape into dynamic typing is particularly useful
in simplifying the incorporation of external JavaScript APIs and
libraries. Sometimes, the user might not want to build a statically
typed API for each external JavaScript library. In addition, for
some library, it might be awkward to come up with such a statically-typed
interface. In general, we expect users to start with a dynamic API for
an external library, and progressively migrate it to a typed API as
the code matures. Consider again the example introduced in the typed
API section:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].getContext()
\end{lstlisting}

In a fully dynamic scenario, we declare the DOM API simply as:
\begin{lstlisting}
trait Dom extends JSDynamic {
  val document: DynamicRep
}
\end{lstlisting}
The \code{as[Canvas]} cast is not necessary in this dynamically typed
setting:
\begin{lstlisting}[escapechar=+]
val context = document.getElementById("canvas").getContext()
\end{lstlisting}

As a first step towards statically typing the API, we declare the type \code{Element}:
\begin{lstlisting}
trait Dom extends JSProxyBase with JSDynamic {
  val document: Rep[Element]
  trait Element
  trait ElementOps {
    def getElementById(id: Rep[String]): DynamicRep
  }
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
}
\end{lstlisting}
Since the method \code{getElementById} returns a \code{DynamicRep},
only the emphasized part of the expression is statically typed:
\begin{lstlisting}[escapechar=+]
val context = +{\it document.getElementById("canvas")}+.getContext()
\end{lstlisting}

@@

First-class functions are widely used in JavaScript. One particular
common case is for callback-driven programming. Therefore, staged
functions are important to interface with existing libraries. They
will also play a crucial role in section~\ref{sec:cps}, where we
abstract over callback-driven programming.

@@

As for implementation, notice that the type of a \code{new JSLiteral
  \{...\}} expression is not \code{JSLiteral \{...\}} but
\code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
Scala-Virtualized compiler. \code{JSLiteral} inherits from a special marker
trait, which indicates its \code{new} expressions should be
reified. So the \code{new} expression is turned into a method call,
with information about all the field definitions. A complication is
that a field definition might reference another field being defined
(such as \code{b} being defined in terms of \code{a} in the example
above). So each definition is represented by its name and a {\it
  function} which takes a self type for the object literal. These
definition functions are evaluated in an order which allows the self
references to be resolved.

@@

Note that JavaScript version takes a callback as second argument that will be used to process the fetched tweets. We provide the rest of the logic that iterates over an array of users and makes Ajax requests (listings~\ref{code:js:twitter_example}~\&~\ref{code:twitter_example}).

@@

\subsection{CPS and Staging}
Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
\begin{lstlisting}
def sleep(delay: Rep[Int]) = shift { k: (Rep[Unit]=>Rep[Unit]) =>
  window.setTimeout(fun(k), delay)
}
\end{lstlisting}

@@

\subsection{CPS for Interruptible Traversals}
We need to be able to suspend our execution while traversing an array in order to implement functionality from listing~\ref{code:twitter_example}. Let's consider a simplified example where we want to iterate over an array and sleep during each iteration. We present both code written in JavaScript and our DSL that achieves that (listings~\ref{code:js:sleep_iter}~\&~\ref{code:sleep_iter}).
\begin{figure}
\lstset{basicstyle=\ttfamily\scriptsize}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\begin{lstlisting}[caption=JavaScript,label=code:js:sleep_iter]
var xs = [1, 2, 3]
var i = 0
var msg = null
function f1() {
  if (i < xs.length) {
    window.setTimeout(f2, xs[i]*1000)
    msg = xs[i]
    i++
  }
}
function f2() {
  console.log(msg)
  f1()
}
f1()
\end{lstlisting}
\begin{lstlisting}[escapechar=|,caption=DSL,label=code:sleep_iter]
val xs = array(1, 2, 3)
// shorthand for xs.suspendable.foreach
for (x <- xs.suspendable) {
  sleep(x * 1000)
  console.log(String.valueOf(x))
}
log("done")

@@

Both programs, when executed, will print the following to the JavaScript console:
\begin{lstlisting}
//pause for 1s
1
//pause for 2s
2
//pause for 3s
3
done
\end{lstlisting}

@@

The Ajax component completes the functionality required to run the Twitter example with one caveat:
The outer loop in listing~\ref{code:twitter_example} uses \code{parSuspendable} to traverse arrays instead of the \code{suspendable} traversal variant we have defined in listing~\ref{code:suspendable_foreach}.

@@

A cell object allows us to track dependencies between values. Whenever the \code{get} method is called and the value is not in the cell yet, the continuation will be added to a queue so it can be suspended until the value arrives. The \code{set} method takes care of resuming queued continuations. We expose \code{Cell} as an external library using our typed API mechanism and we use it for implementing an abstraction for futures.

@@

We've implemented our DSL in Scala.\footnote{The code can be found at
  \url{http://github.com/namin/lms-sandbox} We used our DSL to
develop a few web applications, which are simple but not
trivial. First, we implemented a few drawing examples like the
snowflakes of figure~\ref{fig:snowflakes}. We extended the Twitter
example from section~\ref{sec:cps}, which presents the latest tweets
from selected users in an interactive way. In order to do so, we
incorporated a useful subset of the DOM API and jQuery library using our
typed APIs.

\begin{wrapfigure}{r}{5cm}
\includegraphics[scale=0.2]{drawing}
\end{wrapfigure}
Finally, we developed a collaborative drawing application, which
includes a server-side component (implemented using the Jetty web
server). We use web sockets to communicate between the server and
clients. Each client transmits drawing commands to the server, which
broadcasts them to all the clients. When a new client joins, the
server sends the complete drawing history to the new client, and the
client reconstructs the image by playing the commands. A very
simple improvement is to make the server execute the drawing
commands as well, and keep an up-to-date bitmap of the drawing -- this
can easily be achieved by using the trivial embedding described in
section~\ref{sec:trivial-embedding}. New clients then just obtain the
bitmap instead of replaying the history, which is potentially large.
